---
title: "Course Activities"
author: "Marc A.T. Teunis"
date: "`r Sys.time()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Course Overview (this document)

 - The course overview provides a summary of the course. It is a document that is fluid throughout the course. It is the current document before you. I will make changes to this as we go along to personalize the course to your needs as much as possible.
  - The course workflow will be addressed
  - If applicable the Course Exam will be addressed (Bachelor Course Only)
  - We will get to know some of the conventions used in the course materials to aid readibility and completion of the exercises

## Labs

 - Each day (lesson) consists of 1 or several labs
 - Is a combination of direct instruction and introduction to a new topic  
 - Each topic (lab) is accompanied by exercises. 
 - Some of the exercise will be addressed during instruction. 
 - Some exercises will have to be prepared outside of class. 
 - Each consecutive lesson will follow up on the previous exercises. 
 - During the follow up solution for the problems that were encountered during the preparation of the exercises will be provided 
 - **There is always room for questions!**

## Cases

 - During each lesson (not lesson 1 ;-), I will ask for a case from the participants
 - Some problem you are working on that is relevant for the topics in that day's class
 - If possible I will prepare a short demo, workflow or provide some tips on how to tackle the case problem
 - I will ask for your input on the matter
 
## Lab 1a: Introduction to RStudio 
Chapter \@ref(lab1aintrorstudio)

 1. Getting the course materials from github.com
 1. Creating objects in R
 1. RStudio introduction

## Lab 1b: Introduction to R  
Chapter \@ref(lab1bintror)

 1. Functional and object oriented programming
 1. Functions
 1. Object Class; vectors, dataframes, lists
 1. Vector types
 1. Getting help
 1. Plots
 1. Data examples

## Lab 1c: Dataframes, lists and matrices  
Chapter \@ref(lab1crecursivevectors)

 1. Dataframes
 1. Lists
 1. Matrices

## Lab 2a: Visualize & Explore Data
Chapter \@ref(lab2viz)

 1. Visualizations demo
 1. Build in datasets 
 1. Using the grammar of graphics with `{ggplot2}`
 1. Plot types
 1. `geom_...` and `aes()`
 1. Adding dimensions to a plot 
 1. Solving overplotting
 1. Plot annotation and labels
 1. Saving plots
 1. Examples of `{shiny}` apps

## Lab 2b: RMarkdown

 1. Literate programming
 1. Markdown syntax
 1. Code chunks
 1. Chunk options
 1. Parameters

## Lab 3a: Data Wrangling  
Chapter \@ref(lab3awrangling)

 1. Using `{dplyr}` for data wrangling
 1. The `{dplyr}` verbs
 1. Data wrangling in combination with `{ggplot2}`

## Lab 3b: Functional Programming 
Chapter \@ref(lab3bfunctions)

 1. What is a function
 1. Writing function
 1. Function documentation
 1. Loops and `map` family of functions
 1. Scripts and the `source` function
 1. Why write function
 1. Function elements (name/arguments, body, return)
 1. Function documentation, Roxygen comments
 1. Writing error / warning /  messages
 
## Lab 3c: Tidy data
Chapter \@ref(lab3ctidydata)
 
## Lab 4a: Importing data 
Chapter \@ref(lab4aimportingdata)

 1. Paths
 1. Importing data functions
 1. Excel and other 'foreign' formats
 1. Reading from a zipfile/url

## Lab 4b: Getting Open Data  
Chapter \@ref(lab4bopendata)

 1. Open data sources
 1. Finding data
 1. Importing data into R (.csv, .tsv, .txt, .json, .xml, .xls(x)
 1. Using APIs (Twitter, Kaggle, Google, CBS)
 1. Accessing databases

## Lab 5: Exploratory Data Analysis _or_ The PPDAC Cycle
Chapter \@ref(lab5eda)

 1. The process of EDA
 1. Problem-Plan-Data-Analysis-Conclusion (PPDAC) Cycle
 1. Missingness
 1. Distibutions
 1. Finding patterns 
 1. Graph types
 1. Multidimensional data
 1. Principal Components
 
## Lab 6a: Exploring Assumptions & Models  

 1. Why Assumptions?
 1. Distributions (Gaussian)
 1. Regression
 1. Model output ( `{broom}`)
 
## Lab 6b: Statistics with R (resampling & simple regression) 

## Lab 6c: Statistics with R (multiple regression and ANOVA) 

