---
title: "Things to forget or ignore"
author: "Marc A.T. Teunis"
date: "`r Sys.time()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
library(tidyverse)
```

## Demo's and tutorials
 - Are great resources to learn R
 - Sometimes they use the best practices
 - Sometimes they clearly do not
 
## Four things you will have to forget (or ignore)
During this course, I will provide good alternatives

 1. Please never use `setwd()`
 2. Please do not attach anything with the `attach()` function
 3. Please do not use the `=` (equal to) assignment when creating variables
 4. Please do not use the `stringsAsFactors = FALSE` `in read.csv()`, use `readr::read_...()` functions instead
 
## Setting the working directory with setwd() 
The `setwd()` function sets the working dierectory to the current directory. I will show you why not to use this.

Imagine you are working in the Documents folder in Windows
```{r}
library(here)

name <- here::here("Users", "Your Name Here", "Documents")
dir.create(here("Users"))
dir.create(here("Users", "Your Name Here"))
dir.create(here("Users", "Your Name Here", "Documents"))
```

## Let's create a file with some dummy data
```{r}
tibble(a = rnorm(100), b = runif(100)) %>%
        write_csv(path = file.path(name, "dummy.csv"))
```

## Use setwd
Use the code below -- Replace your own name at "Your Name Here"
```{r}
path <- here("Users", "Your Name Here", "Documents")
setwd(path)
read_csv("dummy.csv")

## Look at where we are
getwd()
```
What do you think, will this code run on another computer?

## Alternative to `setwd()`

 - We will learn about some of the alternatives for using `setwd()`
 - The packages `rprojroot`, and the `here` package, together with the function `file.path()` are good alternatives.
 - I will explain how to use them in the labs on RMarkdown and importing data

## `attach()`
In some of the older R code around you will find the frequent use of the `attach()` function. This function puts R objects in the search path.
This means they are always available for the user. Using `attach()` breaks portability of the code. Be explicit and use relative paths.
If you forget to attach or attach multiple times, the code breaks. This makes running code with `attach()` in it fragile

## There are several reasons why NOT to use `attach()`

 - You may have multiple object in your session carrying the same name, it is easy to mix things up
 - When you attach a dataset it attaches the values (in e.g. a dataframe column) as they were when they were attached. Any changes in between are not attached: Let's illustrate this with an example
 
## Assigning variables

To assign a variable in R you can use:

`=` or `<-`

`=` has a special meaning and is reserved for naming an argument in a function. Always use `<-` when you want to assign a value to a variable. 

Remember: `<-` always works for assignment, `=` does not always work for assignment.

## `{readr}` functions
readr functions never convert to factors, except if explicitly told:
```{r, eval=TRUE}

example_tibble <- read_delim(
        here::here("data", "gender.txt"), 
        delim = "/",
        col_types = cols(
          sex = col_factor()
          )
)

example_tibble
## only `sex` is converted to a factor
```
