---
title: "Conventions, Code and Data Management"
author: "Marc A.T. Teunis"
date: "`r Sys.time()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 5, 
                      fig.height = 3,
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Packages
```{r}
library(usethis)
library(devtools)
library(gapminder)
library(nycflights13)
library(tidyverse)
```

## Code Style

 * The slides in the course contain a number of style and syntax conventions.
 * Code styles and syntax can vary
 * Be consistent

## Capitals vs lower case

 * Everything in lower case
 * I use `snake_case`
 * Many people use `camelCase`

```{r}
gapminder %>% names
flights %>% names
```

## Printing a tbl_df
```{r}
table <- tibble("a_is_a_variable" = c(1:5), "b_is_another_variable" = rnorm(5))
table
```

## Variables

 * All lower case
 * Don't use capitals and special characters in file names

"`excel_file_of_important_dataset.xlsx`"

DON'T: "`data file + version 12_11_2017%.xlsx`"

## Packages referrals

Names of R packages are indicated as:

 * mostly: `{dplyr}`
 * sometimes: {dplyr}
 * not often: "dplyr"

## Explicit functions

When we want to be explicit about a certain function in a package:

`dplyr::filter()`

Refers to the `filter()` function in the package `{dplyr}`

## Loose code in the slides
Sometimes a few lines of codes are displayed for demo purpose:

`four_numbers <- c(1:4)`

`multiply <- 4*four_numbers`

`sum(multiply^2)`

## Code blocks

Code blocks in slides are indicated as below:
The result of the code is usually also displayed
```{r, fig.height=2, fig.width=3}
set.seed(1234)
code_that_does_something <- rnorm(1000)
hist(code_that_does_something, main = "rnorm")
```

## Hiding code 

Sometimes it makes sense to hide the code and only display the result. 
This can be done by setting the code chunk options `echo = FALSE`

```{r, echo=FALSE, fig.height=2.5, fig.width=3}
set.seed(1234)
code_that_does_something <- rnorm(1000)
hist(code_that_does_something, main = "rnorm")
```

## Other important chunk options

 * Like hiding the code, we can hide the result of that code and only display the code by setting `results = "hide"`
 * Setting `eval = FALSE` means the chunk is not included in the slide show at all
 * Setting `include = FALSE` means the chunk is run and the results are evaluated, but both the code and the results are not included in the slide-show/document 
 * Controlling settings for the complete document can be done in the `setup*` chunk that you will find in all Rmarkdown files from the course

```{r, setup2, eval = FALSE}
 knitr::opts_chunk$set(echo = TRUE,
                       message = FALSE,
                       warning = FALSE,
                       error = TRUE,
                       fig.width = 5, fig.height = 3)
```
 
 * There are many chunk options. Some we will see in later lessons

## Replacing old, creating new variables
```{r, eval=FALSE}
## replacing lifeExp in years for lifeExp in months
gapminder$lifeExp <- (gapminder$lifeExp*24)

## instead create a new variable in the dataframe
gapminder <- gapminder %>%
  mutate(lifeExp_months = lifeExp * 24)
```

_Mutate new variables on the basis of old one, do not replace the old variable_

## R packages

 * R packages are self-contained pieces of R code that allow easy sharing of R implementations
 * They follow a strict structure
 * They allow for versioning
 * The can be tested (unit tests)
 * They can be distributed (CRAN, Bioconductor, Github, Bitbucket, ...)
 * They can be made to fit multiple Operating Sytems
 
## There are many R packages

```{r, r_packages_graph, echo=FALSE}
source(here::here("/code/r_packages_graph.R"))
pkgs_plot <- plot_packages()
pkgs_plot
```



 



