---
title: "Course intro"
subtitle: "Data anlaysis with R for staRters"
author: "Marc A.T. Teunis, Ph.D."
date: "`r Sys.time()`"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Slides

_"These are the only slides we will see in this course"_

```{r, echo=FALSE}
knitr::include_graphics(
  here::here(
    "images",
    "rstudio_interface.png"
  )
)
```


## Course aim

 - To get to learn the possibilities of R for data analysis
 - To learn why scripting is essential for reproducible science
 - To get involved in the HU Data Science community
 - To get inspired
 
```{r}
knitr::include_graphics(
  here::here(
    "images",
    "Rlogo.png"
  )
)
```
 
 
## Disclaimer
 
 _"Ten minutes of R a day, keeps Excel away (or SPSS for that matter)"_
 
 - There is (much) more ground than we can cover in 3 days
 - There are more materials that I have included than we will be able to tackle
 - Learning R is about commitment and putting in 'the hours'
 - I can point you in the right direction, the rest is up to you
 - Learning how to program is not the same as learning another software tool
 - R is not _another_ software tool
 - If you learn how to program, you (and science) will benefit from it later, so look at it as an investment
 
## Learning how to program {.smaller .build}

 - Google everything, I do too
 - Has a steep learning curve
 - Try a hard stop for using 'other' tools
 - Don't try to be perfect on the first go, these things take time
 - Reiterate and revisit your code, do not rewrite everything
 - "Do one thing at a time" - "Do not repeat yourself"
 - Be aware that there is no turning back...
 - Once you commit, you will never be the same...
 
```{r, echo=FALSE}
knitr::include_graphics(
  here::here(
    "images",
    "stairs_ungol.jpg"
  )
)
```
 
 
## Etiquettes
 
  - We are with a large group
  - Please raise your hand if you want to speak out
  - Please use the chat for questions
  - If you want to bail out, do so quitely
  - If you don't agree, please speak up
  
## Course flow

 - Demo -> Break -> Assignments, repeat (about 1.5 hours per cycle)
 - Assignments, Three rooms: _Quiet_, _Conversation_, _Questions_
 - When working on assignments you can choose a room. I will hop into the questions room 
 - You can hop in and out different rooms, but be quiet in the _Quiet_ room
 - Please come back in time to the _Plenary Room_ for the plenary demo sessions 
 
## Contents

 1. Intro, syntax, objects
 2. Visualization, tidy data
 3. Wrangling, importing data, statistics
  
## Let's get to it!

```{r, echo=FALSE, dpi=60}
knitr::include_graphics(
  here::here(
    "images",
    "Obama-Yes-We-Can.jpg"
    
  )
)
```

